Type theory with single substitutions\thanks{This research was supported by the the HOTT ERC Grant 101170308.}

### 2025-05-16 lfmtp/p.tex
finish

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tel}{\mathsf{Tel}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\Type}{\mathsf{Type}}

@@

We construct another isomorphism between terms depending on a variable and
terms depending on the lifted version of the same variable:
\begin{equation}\label{eq:liftvar}
(\lambda t\ldotp t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t\ldotp t[\p^+][\langle\mk\,\q\rangle])
\end{equation}
To derive the round-trips, we again need admissible equations. We prove
the first round-trip equality in Appendix \ref{app:liftvar}, the other
is analogous.
Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
\begin{alignat*}{10}
  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) \overset{\text{$\Lift$}}{\cong}
  \Tm\,\Gamma\,(\Pi\,A\,B) \overset{\text{$\Pi$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,B \overset{\text{$\Lift$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,(\Lift\,B) \overset{\text{lifted var}}{\cong} \\
  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) \overset{\text{$\Pi$}}{\cong}
  \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
\end{alignat*}
We hope that the above examples demonstrated the need for equations
(\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
section.

@@

In this section we show that if the single substitution calculus has
certain type formers, then a CwF structure is derivable in any model.
The idea is that
\begin{itemize}
\item using $\Sigma$ types we emulate contexts,
\item using functions between these $\Sigma$ types we emulate parallel
  substitutions,
\item using the functions into the universe we emulate dependent types,
\end{itemize}
and then single substitutions are just there to set up $\Sigma$, $\Pi$
and $\U$, and these types are enough to bootstrap the parallel
substitution calculus.

@@

In Appendix \ref{sec:functor}, we show that the roundtrip CwF

>>>>

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tel}{\mathsf{Tel}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\Type}{\mathsf{Type}}
% \NewCommandCopy{\oldGamma}{\Gamma}\renewcommand{\Gamma}{{\mathit{\oldGamma}}}
% \NewCommandCopy{\oldDelta}{\Delta}\renewcommand{\Delta}{{\mathit{\oldDelta}}}
% \NewCommandCopy{\oldTheta}{\Theta}\renewcommand{\Theta}{{\mathit{\oldTheta}}}

@@

We construct another isomorphism between terms depending on a variable and
terms depending on the lifted version of the same variable:
\begin{equation}\label{eq:liftvar}
(\lambda t\ldotp t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t\ldotp t[\p^+][\langle\mk\,\q\rangle])
\end{equation}
To derive the round-trips, we again need admissible equations. We prove
the first round-trip equality in Appendix \ref{app:tt}, the other
is analogous.
Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
\begin{alignat*}{10}
  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) \overset{\text{$\Lift$}}{\cong}
  \Tm\,\Gamma\,(\Pi\,A\,B) \overset{\text{$\Pi$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,B \overset{\text{$\Lift$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,(\Lift\,B) \overset{\text{lifted var}}{\cong} \\
  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) \overset{\text{$\Pi$}}{\cong}
  \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
\end{alignat*}
We hope that the above examples demonstrated the need for equations
(\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
section.

@@

In this section we show that if the single substitution calculus has
certain type formers, then a CwF structure is derivable in any model.
The idea is that (i) using $\Sigma$ types we emulate contexts; (ii)
using functions between these $\Sigma$ types we emulate parallel
substitutions; (iii) using the functions into the universe we emulate
dependent types. Then single substitutions are just there to set up
$\Sigma$, $\Pi$ and $\U$, and these types are enough to bootstrap the
parallel substitution calculus.

@@

In Appendix \ref{app:functor}, we show that the roundtrip CwF
